Global patterns of forest autotrophic carbon fluxes

### 2020-03-30 manuscript/manuscript draft.Rmd
worked on discussion

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine FACF analyzed, C cycling decreases linearly (or exponentially, in the case of $GPP$) with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. The FACF generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (e.g., *$ANPP_{woody}$*, *$BNPP_{fine.root}$*, $R_{auto-root}$; Fig. 2; Table S2). Latitudianal variation in FACF is primarily attributable to temperature-related variables (*H3, H4*), particularly MAT (Figs. 3-4). Water availability is also influential, but generally of secondary importance across the range represented in our database (Figs. 3-4). Temperature seasonality and growing season length are closely correlated with MAT and are strong predictors of FACF (*H4*; Figs. 4e-f, S2, S6-S7), though growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of FACF (*H5*; Fig. S9; Table S4). These findings clarify the big picture of how FACF vary with latitude and climate on a global scale.

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with suggestions that productivity of temperate forests is similar to that of tropical forests (**REFS**). Temperate forests tend to be younger than tropical forests (**REF**), so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/); Potapov et al. 2008. Mapping the World’s Intact Forest Landscapes by Remote Sensing. Ecology and Society 13 (2), 51. ). In addition, because C allocation varies with stand age [@delucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (** Nobby AFM paper: Understanding the controls over forest carbon use efficiency on small spatial scales: Effects of forest disturbance and tree diversity**) [*purpose for respiration/ other compenents (**See Nobby's comment in manuscript-draft_NK.pdf**)].

We show that FACF are broadly consistent in their responses to climate drivers on the global scale (with the exception of some differences in MAT-MAP interactions; Fig. 3), with no major trends in C allocation among the variable pairs tested (Table 1, *H1*; Fig. 2; **Some SI table**). Although variation in allocation has been observed along gradients of elevation [@moser_elevation_2011] and water availability [@newman_above-_2006]--along with non-climatic axes of stand age [@litton_carbon_2007], nutrient availability [@litton_carbon_2007; @gill_belowground_2016], and forest structure [@taylor_greater_2019]--variation in relation to climate is not apparent at the global scale within ForC, which contains the bulk of relevant data. Our conclusion, then, is that hypothesized gradients in allocation along global climate gradients cannot currently be supported for mature forests, although data quantity and standardization is currently insufficent to rule out the possibility that such trends exist.

bbl: remove this sentence? Of particular interest and significance are the relationships amongst $GPP$, net primary productivity ($NPP$ and its components, particularly $ANPP_{woody-stem}$), and respiration ($R_{auto}$ and components). There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there is indication that CUE or $ANPP_{woody-stem}$/$GPP$ increase with latitude (**some SI table**). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$ - latitude $R^2\ge$ 48%) than in subsidiary fluxes (latitude $R^2$ <27% for $BNPP_{fine.root}$, $R_{auto-root}$,$ANPP_{woody-stem}$) (Fig. 2; **some SI table?**). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (bbl: cite e.g. Huntzinger?). However, methodological variation and uncertainty affect all of fluxes considered here--not necessarily any less than the aforementioned, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more dicrect climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, and secondarily by moisture availability (Table 1, *H2-H3*; Figs. 3-4). Because many climate variables co-vary across the latitudinal gradient, because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means, it is not appropriate to attempt to identify any one mean annual climate variable as a mechanistic driver of FACF. However, it remains informative to consider these relationships. We find that temperature-related climate variables ($MAT$, temperature seasonality, ...**LIST**) explain the highest proportion of variability in FACF, and among these, $MAT$ is generally the best predictor--perhaps because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables are extracted from global gridded data products (Table S1**?**). The effects of temperature are modified by moisture availability, with reduced FACF under hot and dry conditions (*i.e.*, high PET, high deficit; Fig. 4c-d) and sometimes under very high precipitaiton (Figs. 3, 4b). Negative effects of very high precipitation on FACF have been observed previously (**REFS**) and are attributable to nutrient and light limitations (**REFS**). Thus, although temperature and water availability jointly and interactively drive global-scale patterns of FACF.

bbl: this seems really interesting and novel, consider expanding the following paragraph

@@

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. We find no support for non-linear trends in mature forest C cycling with respect to latitude or $MAT$, and no distinct trends in C allocation across the global scale (Fig. 2). The implication is that under warmer conditions with similar moisture availability--and within the temperature range to which forest communities are adapated and acclimatized--higher temperatures result in a generalized acceleration of FACF, with no major shifts in C allocation among subsidiary fluxes. This is consistent with observations of continental- to global-scale increases in $GPP$ (*[Li & Xiao 2019](https://www.mdpi.com/2072-4292/11/21/2563)*), $ANPP_{woody stem}$ (*[Brienen et al., 2015](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature14283)*), as well as some C cycle components not considered here--tree mortality (*[Brienen et al., 2015](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature14283)*; **some McDowell Ref-KAT**), soil respiration ([Bond Lamberty & Thompson 2010](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature08930/)), heterotrophic soil respiration ([Bond-Lamberty et al. 2018](https://www-nature-com.smithsonian.idm.oclc.org/articles/s41586-018-0358-x)). Of course, other factors such as rising CO~2~ are also at play, and such increases are not universal (e.g., [Rutishauser et al., 2019](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14833?casa_token=imtB9LD9mL8AAAAA%3AYWgsqNUfRkgs6_dIuDzmwUzL0qVyyxojhbj3jjfRcykBy8ay9T4qMOKPeOPytSGET9muIs8myhQOh3E))... Of course, actual climatic changes will result in very different sets of conditions than represented across geographic gradients in climate, but our analysis clarifies how carbon cycles through contemporary forest ecosystems. As we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change. **[work on this]**

>>>>

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine FACF analyzed, C cycling decreases continually with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. The FACF generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (*e.g.*, $ANPP_{woody}$, $BNPP_{fine.root}$, $R_{auto-root}$; Fig. 2; Table S2). Latitudianal variation in FACF is primarily attributable to temperature-related variables (*H3, H4*), particularly MAT (Figs. 3-4). Water availability is also influential, but generally of secondary importance across the range represented in our database (Figs. 3-4). Temperature seasonality and growing season length are closely correlated with MAT and are strong predictors of FACF (*H4*; Figs. 4e-f, S2, S6-S7), though growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of FACF (*H5*; Fig. S9; Table S4). These findings clarify the big picture of how FACF vary with latitude and climate on a global scale.

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands ([Poulter et al. 2018-DOI:10.1594/PANGAEA.889943](http://pure.iiasa.ac.at/id/eprint/15361/); Potapov et al. 2008. Mapping the World’s Intact Forest Landscapes by Remote Sensing. Ecology and Society 13 (2), 51. ), so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because C allocation varies with stand age [@delucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (** Nobby AFM paper: Understanding the controls over forest carbon use efficiency on small spatial scales: Effects of forest disturbance and tree diversity**) [*purpose for respiration/ other compenents (**See Nobby's comment in manuscript-draft_NK.pdf**)].

**This paragraph needs work--including re-assessment of results, interpretation, writing**
We show that FACF are broadly consistent in their responses to climate drivers on the global scale (with the exception of some differences in MAT-MAP interactions; Fig. 3), with no pronounced trends in C allocation among the variable pairs tested (Figs. 2, S3). Although variation in allocation has been observed along gradients of elevation [@moser_elevation_2011] and water availability [@newman_above-_2006]--along with non-climatic axes of stand age [@litton_carbon_2007], nutrient availability [@litton_carbon_2007; @gill_belowground_2016], and forest structure [@taylor_greater_2019]--little variation in relation to climate is apparent at the global scale within ForC, which contains the bulk of relevant data. Our conclusion, then, is that hypothesized gradients in allocation along global climate gradients cannot currently be supported for mature forests, although data quantity and standardization is currently insufficent to rule out the possibility that such trends exist. 
**bbl: remove following sentence?** Of particular interest and significance are the relationships amongst $GPP$, net primary productivity ($NPP$ and its components, particularly $ANPP_{woody-stem}$), and respiration ($R_{auto}$ and components). There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there is indication that CUE or $ANPP_{woody-stem}$/$GPP$ increase with latitude (Fig. S3). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

One interesting observation was that climate tends to explain more variation in the major fluxes ($GPP$, $NPP$, $R_{auto}$ - latitude $R^2\ge$ 48%) than in subsidiary fluxes (latitude $R^2$ <27% for $BNPP_{fine.root}$, $R_{auto-root}$,$ANPP_{woody-stem}$) (Fig. 2; Table S2). There are two, non-exclusive, potential explanations for this. First, it may be that methodological variation is larger relative to flux magnitude for some of the subsidiary fluxes. Belowground fluxes in particular are difficult to quantify, and measurement methods for the belowground fluxes considered here may be measured through fundamentally different approaches (*e.g.*, minirhizotrons, ingrowth cores, or sequential coring for $BNPP_{root-fine}$; root exclusion, stable isotope tracking, or gas exchange of excised roots for $R_{auto-root}$), and sampling depth is variable and often insufficient to capture the full soil profile. $ANPP_{woody-stem}$, which is also poorly explained by latitude or climate, is more straightforward to measure but is subject to variability introduced by differences such as biomass allometries applied and minimum plant size sampled (**bbl: cite e.g. Huntzinger?**). However, methodological variation and uncertainty affect all of fluxes considered here--not necessarily any less than the aforementioned, and some of the larger fluxes that vary more strongly with respect to climate ($ANPP$, $NPP$) are estimated by summing uncertain component fluxes. Second, differences among variables in the proportion of variation explained by climate may be attributable to more dicrect climatic control over $GPP$ than subsidiary fluxes. That is, subsidiary fluxes may be shaped by climate both indirectly through its influence on $GPP$ and respiration and directly through any climatic influence on C allocation, as well as many other local- and regional-scale factors (**REFS**).

The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, and secondarily by moisture availability (Table 1, *H2-H3*; Figs. 3-4). Because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means, it is not appropriate to attempt to identify any one mean annual climate variable as a mechanistic driver of FACF. However, it remains informative to consider these relationships. We find that temperature-related climate variables ($MAT$, temperature seasonality, ...**LIST**) explain the highest proportion of variability in FACF, and among these, $MAT$ is generally the best predictor--perhaps because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables are extracted from global gridded data products (Table S1). The effects of temperature are modified by moisture availability, with reduced FACF under hot and dry conditions (*i.e.*, high PET, high deficit; Fig. 4c-d) and sometimes under very high precipitaiton (Figs. 3, 4b). Negative effects of very high precipitation on FACF have been observed previously (**REFS**) and are attributable to nutrient and light limitations (**REFS**). Thus, although temperature and water availability jointly and interactively drive global-scale patterns of FACF.

@@

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ (*[Li & Xiao 2019](https://www.mdpi.com/2072-4292/11/21/2563)*) and $ANPP_{woody stem}$ (*[Brienen et al., 2015](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature14283)*; *[Hubau et al. 2020](https://doi.org/10.1038/s41586-020-2035-0)*), along with some C cycle components not considered here--tree mortality (*[Brienen et al., 2015](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature14283)*; *[McDowell et al. 2018](https://nph.onlinelibrary.wiley.com/doi/full/10.1111/nph.15027 )*), soil respiration ([Bond Lamberty & Thompson 2010](https://www-nature-com.smithsonian.idm.oclc.org/articles/nature08930/)), and heterotrophic soil respiration ([Bond-Lamberty et al. 2018](https://www-nature-com.smithsonian.idm.oclc.org/articles/s41586-018-0358-x)). However, increasing C flux rates are not universal (e.g., *[Rutishauser et al., 2019](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14833?casa_token=imtB9LD9mL8AAAAA%3AYWgsqNUfRkgs6_dIuDzmwUzL0qVyyxojhbj3jjfRcykBy8ay9T4qMOKPeOPytSGET9muIs8myhQOh3E)*, **MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of cliamte, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.
